Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-08-31 WSSSPE3/report/wg_main_eng_design.tex
changing subsections of 5 to not be first person

The group discussion focused on identifying existing principles of software
engineering design that could be adopted by the computational science and
engineering communities.

>>>>

This group's discussion focused on identifying existing principles of software
engineering design that could be adopted by the computational science and
engineering communities.

### 2016-08-31 WSSSPE3/report/wg_main_funding_spec_expert.tex
changing subsections of 5 to not be first person

The solution we envision entails creating an environment where software
developers are a stable part of a research team. Such an environment mitigates
the risk of losing a key developer at a critical moment in a projects lifetime,
and provides the benefits of building a store of institutional knowledge about
specific projects as well as about software development for today's research.
Our vision is to find a way to promote a university/research institute
environment where software developers are stable components of research project
teams.

One strategy to promote stability is implementing a mechanism for developers to
obtain academic credit for software development work. With such a mechanism in
place, traditional academic funding models and career tracks could properly
sustain individuals for whom software development is their primary contribution
to research. A contributing factor to the problem with the current academic
reward system is the devastating effect on an academic publication record
resulting from time in industry; such postings often develop exactly the skills
that research software engineers need, yet returns to university positions
following an industry role are penalized by the current structures. Retention of
senior developers is hard, because these people are high in demand by the
economy. However, people who have a PhD in science and enter industry, may
desire to return for diverse reasons, and should be welcomed back.

While developing new mechanisms in the current academic reward system is a
worthy aspirational goal, such a dramatic change in this structure does not seem
likely in a time scale relevant to this working group. Accordingly, our working
party sought alternative solutions that may be achievable within the context of
existing academic structures. The group felt that developing dedicated research
software engineering roles within the university, and finding stable funding for
those individuals is the most promising mechanism for creating a stable software
development staff.

Measures of impact and success for research programming groups, as well as for
individual research software engineers, will be required in order to make the
case to the university for continued funding. Research software engineers will
not be measured by publications, we hope, but by other metrics. Middle-author
publications are common for RSEs. Most RSEs welcome co-authorship on papers
where the PIs think that the contribution deserves it.

@@

One solution is to fund dedicated software engineering roles for major research
software projects at national laboratories or other non-educational
institutions. This solution is in place and working well for many well-used
scientific codebases. However, this strategy has limited application, as much of
the body of software is created and maintained in research universities.
Therefore, we argue that research institutions should develop hybrid
academic-technical tracks for this capacity, where employees in this track work
with more than one PI, rather than the traditional RA role within a single
group. This could be coordinated centrally, as a core facility, perhaps within
research computing organizations which have traditionally supported university
cyberinfrastructure, library organizations, or research offices. Alternatively,
these groups could be organizationally closer to research groups, sitting within
academic departments. The most effective model will vary from institution to
institution, but the mandate and ways of working should be similar.

Having convinced ourselves that this would be a positive innovation, we were
then faced with the specific question of how to fund the initiation of this
activity. A self-sustaining research software group will support itself through
collaborations with PIs in the normal grant process, with PIs choosing to fund
some amount of research software engineering effort through grants in the usual
way. However, to bootstrap such a function to a level where it has sufficient
reputation and client base to be self-sustaining will generally require seed
investment.

This might come from universities themselves (this was the model that led to the
creation of the group in University College London), but more likely, seed
funding needs to come from research councils (as with the Research Software
Engineering Fellowship provided by the UK Engineering and Physical Sciences
Research Council). We therefore recommend that funding organizations consider
how they might provide such seed funding.

@@

We finally noted that the institutional case for such groups is made easier by
having successful examples to point to. In the UK, a collective effort to
identify the research software engineering community, with individuals clearly
stating ``I am a research software engineer,'' has been important to the
campaign. It will be useful to the global effort to similarly identify emerging
research software organizations, and also, importantly, to identify
longer-running research software groups, which have in some cases had a long
running \emph{sui-generis} existence, but which now can be identified as part of
a wider solution. There remains the problem of how to ``sell'' the value of this
investment to investigators within a university. This is an issue best addressed
by the individual organizations that embark on the plan.

@@

The first step in moving this strategy forward is to gather a list of groups
that self-identify as research software engineering groups, and to reach out to
other organizations to see if there may be a widespread community of RSEs who do
not identify themselves as such at this time. We will collect information about
the organizational models under which these groups function, and how they are
funded. For example, how many research universities currently fund people in the
RSE track, whether they bear the RSE moniker or not. Are these developers paid
by the university or through a program supported by research grants/individual
PIs? How did they bootstrap the developer track to get this started? How
successful is the university in getting investigators to pay for fractional
RSEs? We will author a report describing our findings, should funding be
available to conduct the investigation.

>>>>

The solution this group envisions entails creating an environment where software
developers are a stable part of a research team. Such an environment mitigates
the risk of losing a key developer at a critical moment in a projects lifetime,
and provides the benefits of building a store of institutional knowledge about
specific projects as well as about software development for today's research.
The group's vision is to find a way to promote a university/research institute
environment where software developers are stable components of research project
teams.

One strategy to promote stability is implementing a mechanism for developers to
obtain academic credit for software development work (see \S\ref{sec:software-credit}.) With such a mechanism in
place, traditional academic funding models and career tracks could properly
sustain individuals for whom software development is their primary contribution
to research. A contributing factor to the problem with the current academic
reward system is the devastating effect on an academic publication record
resulting from time in industry; such postings often develop exactly the skills
that research software engineers need, yet returns to university positions
following an industry role are penalized by the current structures. Retention of
senior developers is hard, because these people are high in demand by the
economy. However, people who have a PhD in science and enter industry, may
desire to return for diverse reasons, and should be welcomed back.

While developing new mechanisms in the current academic reward system is a
worthy aspirational goal, such a dramatic change in this structure does not seem
likely in a time scale relevant to this working group. Accordingly, the working
group sought alternative solutions that may be achievable within the context of
existing academic structures. The group felt that developing dedicated research
software engineering roles within the university and finding stable funding for
those individuals is the most promising mechanism for creating a stable software
development staff.

Measures of impact and success for research programming groups, as well as for
individual research software engineers, will be required in order to make the
case to the university for continued funding. Research software engineers will
hopefully not be measured by publications, but by other metrics. Middle-author
publications are common for RSEs. Most RSEs welcome co-authorship on papers
when the PIs think that the contribution deserves it.

@@

One solution is to fund dedicated software engineering roles for major research
software projects at national laboratories or other non-educational
institutions. This solution is in place and working well for many well-used
scientific codebases. However, this strategy has limited application, as much of
the body of software is created and maintained in research universities.
Therefore, the group argues that research institutions should develop hybrid
academic-technical tracks for this capacity, where employees in this track work
with more than one PI, rather than the traditional RA role within a single
group. This could be coordinated centrally, as a core facility, perhaps within
research computing organizations which have traditionally supported university
cyberinfrastructure, library organizations, or research offices. Alternatively,
these groups could be organizationally closer to research groups, sitting within
academic departments. The most effective model will vary from institution to
institution, but the mandate and ways of working should be similar.

Having convinced themselves that this would be a positive innovation, the group members were
then faced with the specific question of how to fund the initiation of this
activity. A self-sustaining research software group will support itself through
collaborations with PIs in the normal grant process, with PIs choosing to fund
some amount of research software engineering effort through grants in the usual
way. However, to bootstrap such a function to a level where it has sufficient
reputation and client base to be self-sustaining will generally require seed
investment.

This might come from universities themselves (this was the model that led to the
creation of the group in University College London), but more likely, seed
funding needs to come from research councils or other funding bodies (as with the Research Software
Engineering Fellowship provided by the UK Engineering and Physical Sciences
Research Council). The group therefore recommends that funding organizations consider
how they might provide such seed funding.

@@

The working group finally noted that the institutional case for such groups is made easier by
having successful examples to point to. In the UK, a collective effort to
identify the research software engineering community, with individuals clearly
stating ``I am a research software engineer,'' has been important to the
campaign. It will be useful to the global effort to similarly identify emerging
research software organizations, and also, importantly, to identify
longer-running research software groups, which have in some cases had a long
running \emph{sui-generis} existence, but which now can be identified as part of
a wider solution. There remains the problem of how to ``sell'' the value of this
investment to investigators within a university. This is an issue best addressed
by the individual organizations that embark on the plan.

@@

The first step in moving this strategy forward is to gather a list of groups
that self-identify as research software engineering groups, and to reach out to
other organizations to see if there may be a widespread community of RSEs who do
not identify themselves as such at this time. This working group will collect information about
the organizational models under which these groups function, and how they are
funded. For example, how many research universities currently fund people in the
RSE track, whether they bear the RSE moniker or not. Are these developers paid
by the university or through a program supported by research grants/individual
PIs? How did they bootstrap the developer track to get this started? How
successful is the university in getting investigators to pay for fractional
RSEs? The group will author a report describing their findings, should funding be
available to conduct the investigation.

### 2016-08-31 WSSSPE3/report/wg_main_industry_interaction.tex
minor changes
changing subsections of 5 to not be first person

At WSSSPE3 our group was interested in better understanding successful pathways
for scientific software to ``transition'' from grant-funded research projects to
industry sponsorship. (This may be an initially awkward phrase---some software
projects will begin their life being sponsored by industry, or result in
collaboration between industry and academia. In such cases, there is still a
need to understand how IP and how maintenance of the software is sustained over
time.)

@@

Most previous research and discussion of industry and academic collaboration,
sharing, and funding of research software has focused on the impact of such
arrangements. Examples of these types of reports are as follows:

@@

Although sustainability transitions are often studied under the broad umbrella
of ``technology transfer,'' we believe there are likely to be a number of
different ways in which a pathway from initial production to long-term
maintenance and secure funding is achieved. In short, industry sponsorship
and/or direct participation is an important aspect of sustaining scientific
software, but our current understanding of these transitions focuses narrowly on
commercial successes or failures of those collaborations.

@@

Plans for carrying forward are currently unclear---this project would require
sustained attention and effort from our team, and at least some amount of
funding in order for us to be involved for extended periods of time.

The broad goals that we would like to accomplish are:

>>>>

At WSSSPE3, this working group was interested in better understanding successful pathways
for scientific software to ``transition'' from grant-funded research projects to
industry sponsorship. (This may be an initially awkward phrase---some software
projects will begin their life being sponsored by industry, or result in
collaboration between industry and academia. In such cases, there is still a
need to understand how IP and how maintenance of the software is sustained over
time.)

@@

Most previous research and discussion of industry and academic collaboration,
sharing, and funding of research software has focused on the impact of such
arrangements. Examples of these types of reports are:

@@

Although sustainability transitions are often studied under the broad umbrella
of ``technology transfer,'' the group believes there are likely to be a number of
different ways in which a pathway from initial production to long-term
maintenance and secure funding is achieved. In short, industry sponsorship
and/or direct participation is an important aspect of sustaining scientific
software, but our current understanding of these transitions focuses narrowly on
commercial successes or failures of those collaborations.

@@

Plans for carrying forward are currently unclear---this project would require
sustained attention and effort from the group members, and at least some amount of
funding in order for those members to be involved for extended periods of time.

The broad goals that the group would like to accomplish are:

### 2016-08-31 WSSSPE3/report/wg_main_metrics.tex
changing subsections of 5 to not be first person

The group discussion began by agreeing on the common purpose of creating a set
of guidance giving examples of specific metrics for the success of scientific
software in use, why they were chosen, what they are useful to measure, and any
challenges and pitfalls; then publish this as a white paper. The group discussed
many questions related to useful metrics for scientific software including
addressing if there is a common set of metrics that can be filtered in some way,
can metrics be fit into a common template, which metrics would be the most
useful for each stakeholder, which metrics are the most helpful and how would we
assess this, how are metrics monitored, and many more. A more complete bulleted
list of these questions can be found in Appendix H. Next, a roadmap for how to
proceed was discussed including creating a set of milestones and tasks. The idea
was put forth for the group to interact with the organizing committee of the
2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in
order to send a software metrics survey to all SI2 and related awardees as a
targeted and relevant set of stakeholders. The five solicitations for software
elements released under the NSF SI2 program all included metrics as a required
component with submitters requested to include {\it ``a list of tangible
metrics, with end user involvement, to be used to measure the success of the
software element developed, \dots''}. These metrics are then reported as part of
annual reports to NSF by the projects. Although neither the proposal text
describing the metrics nor the reported metric results are publicly available,
there is reason to believe that the community will be willing to provide this
information through a survey mechanism. This survey would be created by one of
the student group members. Similarly, it was suggested that a software metrics
survey be sent to the UK SFTF (Software For The Future, led by the Engineering
and Physical Sciences Research Council) and TRDF (Tools and Resources
Development Fund, led by the Biotechnology and Biological Sciences Research
Council) software projects to ask them what metrics would be useful to report.
The remainder of the discussion focused mainly on the creation of a white paper
on this topic. This resulted in a paper outline and writing assignments with the
goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of
Electrical and Electronic Engineers Computing in Science and Engineering
magazine), or JORS (Journal of Open Research Software). More information about
the group discussion is available in Appendix~\ref{sec:appendix_metrics}.

>>>>

The group discussion began by agreeing on the common purpose of creating a set
of guidance giving examples of specific metrics for the success of scientific
software in use, why they were chosen, what they are useful to measure, and any
challenges and pitfalls; then publish this as a white paper. The group discussed
many questions related to useful metrics for scientific software including
addressing if there is a common set of metrics that can be filtered in some way,
can metrics be fit into a common template, which metrics would be the most
useful for each stakeholder, which metrics are the most helpful and how would we
assess this, how are metrics monitored, and many more. A more complete bulleted
list of these questions can be found in Appendix~\ref{sec:appendix_metrics}. Next, a roadmap for how to
proceed was discussed, including creating a set of milestones and tasks. The idea
was put forth for the group to interact with the organizing committee of the
2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in
order to send a software metrics survey to all SI2 and related awardees as a
targeted and relevant set of stakeholders. The five solicitations for software
elements released under the NSF SI2 program all included metrics as a required
component with submitters requested to include {\it ``a list of tangible
metrics, with end user involvement, to be used to measure the success of the
software element developed, \dots''}. These metrics are then reported as part of
annual reports to NSF by the projects. Although neither the proposal text
describing the metrics nor the reported metric results are publicly available,
there is reason to believe that the community will be willing to provide this
information through a survey mechanism. This survey would be created by one of
the student group members. Similarly, it was suggested that a software metrics
survey be sent to the UK SFTF (Software For The Future, led by the Engineering
and Physical Sciences Research Council) and TRDF (Tools and Resources
Development Fund, led by the Biotechnology and Biological Sciences Research
Council) software projects to ask them what metrics would be useful to report.
The remainder of the discussion focused mainly on the creation of a white paper
on this topic. This resulted in a paper outline and writing assignments with the
goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of
Electrical and Electronic Engineers Computing in Science and Engineering
magazine), or JORS (Journal of Open Research Software). More information about
the group discussion is available in Appendix~\ref{sec:appendix_metrics}.

### 2016-08-31 WSSSPE3/report/wg_main_software_credit.tex
changing subsections of 5 to not be first person

Since near-term efforts will be shifting to the FORCE11-SCWG, we direct
interested readers to that group's existing landing page\footnote{FORCE11-SCWG
landing page,
\url{https://www.force11.org/group/software-citation-working-group}} and GitHub
repository\footnote{FORCE11-SCWG GitHub page,
\url{https://github.com/force11/force11-scwg}}.

>>>>

Since near-term efforts will be shifting to the FORCE11-SCWG, 
interested readers should go to that group's existing landing page\footnote{FORCE11-SCWG
landing page,
\url{https://www.force11.org/group/software-citation-working-group}} and GitHub
repository\footnote{FORCE11-SCWG GitHub page,
\url{https://github.com/force11/force11-scwg}}.

### 2016-08-31 WSSSPE3/report/wg_main_user_community.tex
changing subsections of 5 to not be first person

There are a number of activities already in progress that are targeted at improving
the user community for open-source software, including Mozilla Science's ``Working Open Project
Guide''~\cite{working-open-wssspe3} and
``UK Collaborative Computational Projects'' (CCP: http://www.ccp.ac.uk), or 
 books such as ``Art of Community'' by Jono Bacon~\cite{art-of-community}.

@@

Everyone agreed on a few points: software must not only offer value, but there
must be some support for users; and funding can help pay for that support, in
addition to the usual funding for software development. Openness is generally 
a virtue. An evangelist, either in the form of a single person or some
domain-specific group of users, is often the key factor.

>>>>

There are a number of activities already in progress that are targeted at improving
the user community for open-source software, including Mozilla Science's ``Working Open Project
Guide''~\cite{working-open-wssspe3} and
``UK Collaborative Computational Projects'' (CCP)\footnote{\url{http://www.ccp.ac.uk)}}, or 
 books such as ``Art of Community'' by Jono Bacon~\cite{art-of-community}.

@@

All the group members agreed on a few points: software must not only offer value, but there
must be some support for users; and funding can help pay for that support, in
addition to the usual funding for software development. Openness is generally 
a virtue. An evangelist, either in the form of a single person or some
domain-specific group of users, is often the key factor.
